---
title: "Experiment005: DC Bikeshare data"
subtitle: "EDA on Sep23 ride data"
author: "Joe Skufca
date:  "2023-11-06"
output: html_notebook:
  toc: yes
---

This experiment continues are class work on the NYCflights data, with a bit more looking at delays.

We will continue to refer to (https://r4ds.hadley.nz/data-transform).


# Packages

Data is from https://s3.amazonaws.com/capitalbikeshare-data/index.html .  
The other packages will support our analysis. 

```{r}
library(tidyverse)
library(janitor)
library(here)

```
### Other project defaults

For consistency:

```{r}
theme_set(theme_minimal())
```





# Read the data

```{r}
df1=read_csv(here("data_raw","202309-capitalbikeshare-tripdata.zip"))
```

# Rides vs time


### Change to long table and computer ridership

To track active riders over time, we'll break each observation into a 

* "ride start", which increases active riders by one
* "ride end" which decreases active riders by one



```{r}
df2s=df1 %>% 
  select(rideable_type,member_casual,
                    contains("start"),ride_id) %>% 
  mutate(start_stop="start") %>%
  rename(t=started_at,
         station_name=start_station_name,
         station_id=start_station_id,
         lat=start_lat,
         lng=start_lng)
df2e=df1 %>% 
  select(ride_id,rideable_type,member_casual,
                    contains("end")) %>%
  mutate(start_stop="stop") %>%
  rename(t=ended_at,
         station_name=end_station_name,
         station_id=end_station_id,
         lat=end_lat,
         lng=end_lng)

df2=bind_rows(df2s,df2e) %>%
  arrange(t) %>%
  mutate(rider_delta=(start_stop=="start")*2-1) %>% #change in ridership 
  mutate(riders=cumsum(rider_delta)) %>%
  relocate(riders,.after=t)


```


### EDA plots of ridership 

#### Full month

```{r}
df2 %>% 
  ggplot(aes(t,riders)) +
  geom_line()
```

#### One day:


```{r}
df2 %>% 
  filter(day(t)==18) %>%
  ggplot(aes(t,riders)) +
  geom_line() +
  ggtitle("Riders on 18Sep")
```

### Can I see ridership with panels:

I want to understand the relationship to weekly patterns:




```{r}
df2 %>% filter(month(t)==9) %>%
  ggplot(aes(t,riders)) +
  geom_line() + 
  facet_wrap(~mday(t),scales = "free_x",ncol=7 )
```
Let's look by weekday



```{r}
df2 %>% filter(month(t)==9) %>%
  mutate(epi=epiweek(t),wd=wday(t,label=TRUE,week_start = 7)) %>%
  ggplot(aes(hms::as_hms(t),riders,color=as.factor(epi))) +
  geom_line(alpha=.7) + 
  facet_wrap(~wd,scales = "free_x",ncol=7 )
```

### Is the pattern similar between electric and classic bikes?

We will need to re-compute the cumsum by group (in addition to overall):


```{r}
df3=bind_rows(df2s,df2e) %>%
  arrange(t) %>%
  mutate(rider_delta=(start_stop=="start")*2-1) %>% #change in ridership 
  mutate(riders=cumsum(rider_delta)) %>%
  mutate(type_riders=cumsum(rider_delta),.by=rideable_type) %>%
  relocate(riders,type_riders,.after=t)
```


```{r}
df3 %>% filter(month(t)==9) %>%
  ggplot(aes(t,type_riders,color=rideable_type)) +
  geom_line() + 
  facet_wrap(~mday(t),scales = "free_x",ncol=7 )
```
Some questions?

https://ride.capitalbikeshare.com/blog/capital-bikeshare-and-dockless-bike-systems-your-top-3-questions-answered



## What is a typical duration?

```{r}
df4=df1 %>% 
  mutate(dur=ended_at-started_at)
```




```{r}
df4 %>% filter(dur<60*45,dur>0) %>%
  ggplot(aes(dur,fill=member_casual)) +
  geom_histogram(alpha=.5)
```



```{r}
df4 %>% filter(dur<60*30,dur>0) %>%
  ggplot(aes(dur,fill=member_casual)) +
  geom_density(alpha=.5)
```



```{r}
df4 %>% filter(dur<60*30,dur>0) %>%
  ggplot(aes(dur,fill=rideable_type)) +
  geom_density(alpha=.5)
```


#### Does duration vary with "when" people ride?


```{r}
df4 %>% 
  filter(dur<60*30,dur>0) %>%
  filter(day(started_at)==18) %>%
  slice_sample(n=10000) %>%
  ggplot(aes(started_at,dur)) +
  geom_point(alpha=.1) 


#+ 
#  facet_wrap(~mday(t),scales = "free_x",ncol=7 )
```

## Let look at ridership by station

```{r}
df5=df2 %>%
  mutate(station_riders=cumsum(rider_delta),.by=station_name)
```

```{r}
df5 %>% tabyl(station_id) %>% arrange(-n)
```



```{r}
df5 %>% 
  filter(station_id %in% c(31005,31258,31221,31623)) %>%
  filter(day(t) %in% c(21,22 )) %>%
  mutate(station_id=as_factor(station_id)) %>%
  ggplot(aes(t,station_riders,color=station_id))+
  geom_line()
```



