---
title: "Replacement Class - Introduction to spatial data"
date:  "2023-11-12"
author: "Coach Skufca"
output: html_notebook
---

This notebook supports an asynchronous lecture meant to introduce you to some basics of spatial data.  The lecture will be provided via Echo360.

I will provide you with a partially completed notebook to assist your work.

This notebooks will focus on "spatial data" and also accessing demographic data in R.

Key topics for today:

* Using `here` package
* Using the `/data_raw/` directory
* Reminder on joins (focus: left join)
* Our Spatial Data
   * Neighborhoods
   * Joining with non-spatial data
   * Census data
   * Joining with spatial data
* ignore html in git

## Packages

Standards:

```{r}
library(knitr)
library(tidyverse)
library(janitor)
library(lubridate) # because we will probably see some dates
library(here) # more easily access files in your project
```

Some additional packages focuses on today's work:

```{r}
library(sf) # working with simple features - geospatial
library(tmap)
library(tidycensus)

```

A link to a book on `tmap`: https://r-tmap.github.io/



## Using the Neighborhood Geospatial Data (using /data)

One way the DC is subdivided is by "neighborhood".

https://opendata.dc.gov/datasets/DCGIS::dc-health-planning-neighborhoods/about


I will use the GeoJSON file.  (Newer, not necessarily better, but ... a single file.  Not smaller, but ... this one is not big.)  



Data is easily readable 
```{r}
neigh=st_read(here("data_raw",
                   "DC_Health_Planning_Neighborhoods.geojson")) %>% clean_names()

class(neigh)
```

```{r}
plot(neigh)
```



## Reminder - Joins

```{r}
df1=tibble(fruit=c("apple","banana","cherry"),cost=c(1.5,1.2,2.25))
df2=tibble(fruit=c("apple","apple","cherry","lemon"),
           desert=c("pie","cobbler","cobbler","cheesecake"),
           cal=c(400,430,500,550))
df1
```
```{r}
df2
```
```{r}
left_join(df1,df2,by="fruit")
```

## Investigating joining spatial and non-spatial data

Read cases information:

```{r message=FALSE, warning=FALSE}
df_c=read_csv(here("data_raw",
                   "DC_COVID-19_Total_Positive_Cases_by_Neighborhood.csv")) %>% 
  clean_names() 
```


```{r}
df_cases=df_c %>%
  filter(as_date(date_reported) == "2021-11-17") %>% 
  separate(neighborhood,into=c("code","name"),sep = ":")

```


```{r}
df_cases=df_c %>%
  filter(as_date(date_reported) == "2021-11-17") %>% 
  separate(neighborhood,into=c("code","name"),sep = ":") %>%
  mutate(code=case_when(code=="N35" ~"N0",
                        TRUE ~ code)) %>%
  select(-date_reported)
  
  
```


## Regular joining (of dataframes)

```{r}
neigh2=left_join(neigh,df_cases,by=c("code")) 

tmap_mode("view")

tm_shape(neigh2) +tm_polygons("total_positives",alpha=.5)
```


## Joining with other spatial data

Let's get some data using `tidycensus`.  Need an API key   https://api.census.gov/data/key_signup.html


```{r eval=FALSE, include=FALSE}
census_api_key("YOUR KEY HERE")

```
 
 
 What data is available --- and what is the variable name?
 
```{r}
#what variables
v20 = load_variables(2018,"acs5")
# median_family_income="	B06011_001" 
# all "B00001_001"	
#black "B02009_001"
```


Get some data:

```{r}
df_cencus=get_acs(geography = "tract",
                  variables=c("median_inc"="B06011_001",
                              "pop"="B01001_001",
                              "pop_black"="B02009_001"),
                  state="DC",geometry=TRUE,year=2021) 
```

```{r}
class(df_cencus)
plot(df_cencus)
```
It's in long format.  Let's make it wide.
```{r}
df_cens=df_cencus %>% 
  select(-moe) %>% 
  pivot_wider(names_from = "variable", 
              values_from = "estimate")
  
 

tm_shape(df_cens) +tm_polygons("median_inc",alpha=.5)
```

### How to join

Consider this problem:

```{r}

  tm_shape(neigh2) +tm_borders(col="blue",lwd=5,alpha=.2)+
  tm_shape(df_cens) +tm_borders(col="red",lwd=1,alpha=.3)
```



```{r}

df_j=st_join(df_cens,neigh2)

```

```{r}
df_cens_adj=df_cens %>% st_transform(4326)
```


```{r}
df_j=st_join(df_cens_adj,neigh2,largest=TRUE)
```
Other order?:

```{r}

df_j_rev = st_join(neigh2,df_cens_adj,largest=TRUE)

```

Since we want the geometry for the NEIGHBORHOODS, we need a different work a little harder:

```{r}
df1=df_j %>% select(median_inc,pop,pop_black,code) %>%
  group_by(code) %>%
  summarise(pop_n=sum(pop),
            pop_black_n=sum(pop_black), 
            adj_median_income=sum(pop*median_inc)/pop_n) 

plot(df1)
```

Now that we are aggregating in the right way, we can join.

```{r}
#df2=left_join(neigh2,df1)

df2=left_join(neigh2,df1 %>% st_set_geometry(NULL))

```

```{r}
df2=df2 %>% mutate(black_perc=pop_black_n/pop_n, covid_rate=total_positives/pop_n)
tm_shape(df2)+tm_polygons(c("adj_median_income","covid_rate","black_perc"))
```



```{r}
df2 %>% filter(code!="N0") %>% tm_shape()+tm_polygons(c("adj_median_income","covid_rate","black_perc"),alpha=.4)
```


